---
title: "Assignment 2: Coding Basics"
author: "Cassidy White"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.



## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1. 
sequenceby4<-seq(1,100,4) #creating a sequence of numbers from 1:100, listing every 4th number
sequenceby4
#2. 
mean<-mean(sequenceby4) #calculate the average of the sequence created above
median<-median(sequenceby4) #calculate the median of the sequence created above
mean
median
#3. 
mean>median #R will let me know if the mean is greater than the median by returning TRUE or FALSE
```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}
names<-c("Tyler", "Mike","Janet","Betty")
names
scores<-c(80, 35, 93, 65)
scores
passed<-c(TRUE, FALSE, TRUE, TRUE)
passed
df_names<-as.data.frame(names) #convert names vector to data frame
df_scores<-as.data.frame(scores) #convert scores vector to data frame
df_passed<-as.data.frame(passed) #convert passed vector to data frame
#class(df_names) #check that conversions were successful
#class(df_scores)
#class(df_passed)

df_summary<-cbind(df_names,df_scores,df_passed)
df_summary
colnames(df_summary)[1]<-"Names"
colnames(df_summary)[2]<-"Scores"
colnames(df_summary)[3]<-"Passed"
df_summary
```

9. QUESTION: How is this data frame different from a matrix?

> Answer: This data frame is different from a matrix because different columns have different value types. Some are numeric, some are character, and some are boolean. In a matrix, all rows and columns must have the same value type.

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

```{r}
passorfail<-function(i){ #trying to create a function to print FALSE if the score is less than or equal to 50 and TRUE if otherwise
  if(i <= 50) {FALSE}
  else {TRUE}
}
```

```{r}
p.or.f<-function(i){ #trying another method for creating a function that prints TRUE if a value is greater than or equal to 50 and FALSE if otherwise
  ifelse(i>=50, TRUE, FALSE)
}
```


11. Apply your function to the vector with test scores that you created in number 5. 

```{r}
score_reveiew<-passorfail(scores) #running the function with separate "if" "else" commands
score_reveiew
```

```{r}
score.review<-p.or.f(scores) #running the function with single line "ifelse" command
score.review
```


12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: Only 'ifelse' worked. Using the separate 'if' and 'else' functions can only process one value and so only returns a TRUE or FALSE for the first score in the vector. The 'ifelse' function can process a vector with multiple values.
