---
title: "Assignment 3: Data Exploration"
author: "Cassidy White, Section #1"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration. 

## Directions
1. Change "Student Name, Section #" on line 3 (above) with your name and section number.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "FirstLast_A03_DataExploration.Rmd") prior to submission.

The completed exercise is due on <>.


## Set up your R session

1. Check your working directory, load necessary packages (tidyverse), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. **Be sure to add the `stringsAsFactors = TRUE` parameter to the function when reading in the CSV files.**
```{r,  message = FALSE}
getwd()
library(tidyverse)
library(ggplot2)
Neonics<-read.csv("../Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", stringsAsFactors = TRUE)
Litter<-read.csv("../Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv", stringsAsFactors = TRUE)

view(Neonics)
view(Litter)
```

## Learn about your system

2. The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicologoy of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: If neonicotinoids are applied to crops as an insecticide, they will have an effect on all insects that come into contact with the crop. Therefore, the application of neonicotinoids can unintentionally impact the populations of beneficial insects like pollinators. We might be interested in the ecotoxicology of neonicotinoids on insects to look at relationships between the pesticide's application and declining bee populations or declining crop yields due to lack of pollination. 

3. The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: Litter and woody debris plays a huge role in a variety of ecological processes. It adds organic matter to the soil that in turn provides nutrients to living plants. We may then be interested in litter and woody debris to look at soil quality/soil chemistry and/or plant growth. Woody debris especially plays a role in wildfire by providing fuel for fires to ignite with and spread. We may then be interested in woody debris to identify regions with high fire risk or to correlate woody debris to variables associated with past fires.

4. How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here: 

> Answer: 
*Sampling is conducted at terrestrial NEON sites with woody vegetation >2m tall.
*Frequency and timing of sampling varies by vegetation type and trap type. 
*The locations for sampling plots are randomly selected.



## Obtain basic summaries of your data (Neonics)

5. What are the dimensions of the dataset?

```{r}
dim(Neonics)
```

6. Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?
```{r}
summary(Neonics$Effect)
```

> Answer: The most common effects studied are population, mortality and feeding behavior. These effects might be studied if researchers were looking into the effectiveness of neonicotinoids. Researchers would be interested in understanding how successful the pesticide was at killing insects, reducing their population, and altering their feeding behavior so that they avoided crops. 

7. Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.

```{r}
summary(Neonics$Species.Common.Name)
```

> Answer: The six most commonly studied insects in this dataset are Honey Bee, Parasitic Wasp, Buff Tailed Bumblebee, Carniolan Honey Bee, Bumble Bee, and Italian Honeybee. Most of these species are pollinators and are likely of interest because they are not the target insect for insecticides but are largely being affected by their application. In fact, bee populations have declined at alarming levels and the decreases in their population are likely related to the application of pesticides, hence the frequency of study.

8. Concentrations are always a numeric value. What is the class of Conc.1..Author. in the dataset, and why is it not numeric?

```{r}
class(Neonics$Conc.1..Author.)
```

> Answer: The class of Conc.1..Author is a factor. I'm guessing it is a factor because we used stringAsFactor = TRUE when importing the dataset.



## Explore your data graphically (Neonics)

9. Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year. 
```{r}
ggplot(Neonics)+
  geom_freqpoly(aes(x=Publication.Year))
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.
```{r}
ggplot(Neonics)+
  geom_freqpoly(aes(x=Publication.Year, color = Test.Location))
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The plot has four lines each shadded differently to represent the count of studies conducted by publication year for each type of test location (Field artificial, Field natural, Field undeterminable, and Lab). Lab and Field Natural location types have both increased significantly in the past two decades and comprise the majority of neonicotinoid tests.

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

```{r}
ggplot(Neonics, aes(x=Endpoint))+
  geom_bar()
```

> Answer: The two most common endpoints are NOEL and LOEL. NOEL stands for no observable effect level and represents the highest concentration at which no effect was observed in the test subject. LOEL stands for lowest observable effect level and represents the lowest concentration of a substance at which an effect was observed in the test subject. 
>

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018. 
```{r}
class(Litter$collectDate)
Litter$collectDate<-as.Date(Litter$collectDate)
unique(Litter$collectDate)
```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
unique(Litter$namedLocation)
```

> Answer: Twelve unique plots were sampled at Niwot Ridge. The information obtained from the unique function is different from the summary function in that it returns the unique values from the selected column. The summary function summarizes the statistics of the selected column.

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.
```{r}
ggplot(Litter, aes(x=functionalGroup))+
  geom_bar()
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup. 
```{r}
ggplot(Litter) +
  geom_boxplot(aes(x = functionalGroup, y = dryMass))
ggplot(Litter) +
  geom_violin(aes(x = functionalGroup, y = dryMass), 
              draw_quantiles = c(0.25, 0.5, 0.75))
```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: THe boxplot is the more effective visualization option because there is so little variation in the data, the violin plots are nearly indiscernible. 

What type(s) of litter tend to have the highest biomass at these sites? 

> Answer: Needles, mixed litter, and twigs/branches tend to have the highest biomass at these sites.